---
  output: github_document
---
  
  <!-- README.md is generated from README.Rmd. Please edit that file -->
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  dev = "png",
  out.width = "80%"
)
```

# INLA.ews

<!-- badges: start -->
[![R-CMD-check](https://github.com/eirikmn/INLA.ews/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/eirikmn/INLA.ews/actions/workflows/R-CMD-check.yaml)
[![License: GPL v2](https://img.shields.io/badge/License-GPL_v2-blue.svg)](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)
<!-- badges: end -->
  
  This Repository contains the INLA.ews package for Bayesian detection of early warning signals. 
  
When a dynamical system approaches a bifurcation point the state variable exhibits increased correlation and variance, this is also known as critical slowing down. Our model assumes that the memory coefficient increases linearly as a function of time, and uses the [R-INLA package](https://www.r-inla.org) to obtain the posterior marginal distributions. Of particular interest is the slope parameter which indicate whether or not correlation is increasing. 
  
INLA.ews currently supports time-dependent AR(1) and fGn processes, although the latter is far more computationally costly.

## Installation

You can install the development version of INLA.ews from [GitHub](https://github.com/) with:
  
``` {r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("eirikmn/INLA.ews")
```

## Example

This is a basic example which shows you how to use the package to perform Bayesian analysis of a simulated time series which exhibit early warning signals in the form of increasing correlation. 

```{r example, eval=TRUE, message=FALSE, warning=FALSE}
library(INLA.ews)

# Set seed and parameters
set.seed(0)
n = 1000
sigma = 1
a=0.2
b=0.7/n
time = 1:n
phis = a+b*time

# Simulate time dependent process
data=numeric(n)
data[1] = rnorm(1,mean=0,sd=sigma)
for(i in 2:n){
  data[i] = rnorm(1, mean=phis[i]*data[i-1],sd=sigma)
}

# Run inla.ews
object = inla.ews(data,model="ar1", memory.true=phis)
```

The plot below shows the estimated posterior marginal mean for the lag-one correlation coefficient (blue) in the above example, with 95% credible intervals included (red). The black line represents the 'true' development of the memory.

```{r plot, message=FALSE, eval=TRUE, echo=FALSE}
require(ggplot2)
plot.df = data.frame(time=object$.args$inladata$time,
                           mean=object$results$summary$phi$mean,
                           median=object$results$summary$phi$median,
                           lower=object$results$summary$phi$hpd0.95lower,
                           upper=object$results$summary$phi$hpd0.95upper)
      ylim=c(0,1)
      ylab=expression(paste("lag-one correlation: ",phi))
ggplot(data=plot.df,aes(x=.data$time)) + 
      geom_ribbon(aes(ymin=.data$lower,ymax=.data$upper),color="red",fill="red",alpha=0.3)+
      theme_bw()+ylab(ylab)+xlab("Time")+ylim(ylim)  + geom_line(aes(y=.data$mean),color="blue") + 
        geom_line(data=data.frame(time=object$.args$inladata$time,
                                  true=object$.args$memory.true),
                  aes(y=.data$true),col="black",alpha=0.8)
```

## Attribution

This code is associated and written for an upcoming paper. Feel free to use the code, but please cite the accompanying paper (when it is published).

## License

The code in this repository is made available under the terms of the GNU General Public License (GPL) version 2 License. For details, see LICENSE.md file.

